# Digital elevation data sets

## Sources of data for Digital Terrain Models

On the global scale there are few sources available, including:

-   SRTM90, SRTM30
-   TanDEM-X
-   Copernicus DEM

NASA held the SRTM\index{SRTM} (Shuttle Radar Topography Mission) in 2000 [@vanzylShuttleRadarTopography2001]. The first set of data was released in 2005, however the regions outside the United States were sampled at 3 arc-seconds, which is 1/1200th of a degree of latitude and longitude, or roughly 90 meters on equator [@farrShuttleRadarTopography2007]. In 2014 NASA released brand-new dataset with a resolution of 1 arc-second, or roughly 30 meters. [Earth Explorer](http://earthexplorer.usgs.gov/) provides the data. The accuracy of the data was assessed by @smithAccuracyResolutionShuttle2003 and later by @rodriguezAssessmentSRTMTopographic2005.

TanDEM-X\index{TanDEM-X} (TerraSAR-X add-on for Digital Elevation Measurements) data produced by SAR interferometry was acquired in January 2015 [@kriegerTanDEMXRadarInterferometer2013; @zinkTanDEMXNewGlobal2014]; the production of the global DEM was completed in September 2016. The absolute height error is within about 1&nbsp;m. The TanDEM-X 90m DEM publicly available has a pixel spacing of 3 arc-seconds, similar to SRTM90. [@wesselTanDEMXGroundSegment2018] It covers with 150 Mio sqkm all Earth's landmasses from pole to pole. More information and data available from [EOC Geoservice](https://geoservice.dlr.de/web/dataguide/tdm90/) portal. The assessment of the data was performed by @wesselAccuracyAssessmentGlobal2018 and @grohmannEvaluationTanDEMXDEMs2018 (for part of Brasil). The 12&nbsp;m and 30&nbsp;m global TanDEM-X DEMs can be ordered by scientific users via a proposal submission.

For the Europe, there is a set released by Copernicus Programme named EU-DEM v1.0 (or v1.1) --- it's a digital surface model (DSM)[^1] of European Economic Area countries. It is a hybrid product based on SRTM and ASTER GDEM data fused by a weighted averaging approach. Detailed description of the dataset and links for download are available under [their portal](https://dataspace.copernicus.eu/explore-data/data-collections/copernicus-contributing-missions/collections-description/COP-DEM). Global 30&nbsp;m and 90&nbsp;m datasets are available worldwide with a free license.

[^1]: Digital Surface Model, not Digital Terrain Model!

All of the above models are characterized by a spatial resolution of 30/90 meters. If more accurate data is needed, look for national sites, where you can download (or buy) the dataset. A few examples are covered in @sec-lidar_dtm_data.


## Data processing

[Looking for a DEM elevation model for entire UK - not in tile form](https://opendata.stackexchange.com/questions/20498/looking-for-a-dem-elevation-model-for-entire-uk-not-in-tile-form) is the query on Open Data StackExchange that will be used as an example of extraction of elevation data from DTM/DEM. The initial question was:

> I need to calculate the elevation of 1600 survey plots covering all of the UK - I have these formatted as a point shapefile. I have looked at the OS50 dataset but this is only available as tiles and it would take forever to process each tile individually - my points cover the entire country. Is there free-to-use raster elevation data anywhere that is not split into many tiles?

Let's acquire and process the data. To download the SRTM rasters we can use one of `elevation_xx()` functions from `geodata` package [@geodata].

```{r}
#| label: ele
#| message: false
#| error: false

ele <- geodata::elevation_30s("United Kingdom", path = "data")
terra::plot(ele)
```

`elevation_30s()` function returns an aggregated data with spatial resolution about 1 km (30 arc seconds). The data was prepared based on SRTM 90 m and supplemented with the GTOP30 data for latitudes greater than 60 degrees.

#### Species occurence data

We will use species occurrence data as an example of points for altitude sampling over the UK. Sample occurrence data for _Calystegia x pulchra_ is taken from GBIF database using `rgbif` package [@rgbif1; @rgbif2].

```{r}
#| label: rgbif
#| output: true

occ <- rgbif::occ_data(
  scientificName = "Calystegia pulchra", 
  country = "GB", 
  hasCoordinate = TRUE
  )

occ <- head(occ$data, 20) |>
  sf::st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = terra::crs(ele)) |>
  subset(select = c("key", "scientificName"))
```

Let's add the observations to the plot:

```{r}
#| label: ele_plot
#| echo: true
#| output: true
terra::plot(ele, main = "Sample of C. pulchra distribution in UK")
occ |>
  sf::st_geometry() |>
  plot(pch = 21, bg = "white",  add = TRUE)
```

Having our sample data prepared (which is class of `r class(occ)`) we can add the elevation to the set using `extract()` function form `terra`. It requires an object of class SpatVector.

```{r}
#| label: elevation_extract
#| message: true
occ |>
  dplyr::mutate(ele_30s = terra::extract(ele, terra::vect(geometry))$GBR_elv_msk) |>
  head(5)
```

Points with no value (`elevation = NaN`) may be caused by inaccuracy or rounding of coordinates when data is normalized by GBIF. In fact, if we checked the third point, it would turn out to be at sea.[^2] Elevations are given in meters.

[^2]: There is a [CoordinateCleaner](https://github.com/ropensci/CoordinateCleaner/) package to deal with such cases.

Let's try more accurate data for UK with 3 arcs sec resolution. Function `elevation_3s(lon, lat, ...)` from `geodata` will help to acquire all necessary tiles. To get the `lon` and `lat` parameters we will download the UK boundaries and check the bounding box.

```{r}
#| label: elevation_3s
#| error: true
uk <- geodata::gadm("United Kingdom", path = "data") |>
  terra::project(terra::crs(ele))

bbox <- sf::st_bbox(uk)

for (x in floor(bbox[1]):ceiling(bbox[3])) {
  for (y in floor(bbox[2]):ceiling(bbox[4])) {
    geodata::elevation_3s(lon = x, lat = y, path = "data")
  }
}

```

Hmm... In fact the boundary of UK exceeds the 60`r r2symbols::sym("degree")` N, so, let's modify our loop (we sacrificed the Shetland Islands, sorry):

```{r}
#| label: elevation_60

for (x in floor(bbox[1]):ceiling(bbox[3])) {
  for (y in floor(bbox[2]):60) {
    geodata::elevation_3s(lon = x, lat = y, path = "data")
  }
}
```

Having all files downloaded we can build an virtual raster and work with it:

```{r}
#| label: virtual_srtm
#| warning: false
#| message: false
#| output: false

r <- list.files(path = "data/elevation", pattern = "srtm_.+.tif", full.names = TRUE)
vrt <- terra::vrt(r, "data/elevation_vrt.vrt", overwrite = TRUE) |>
  terra::crop(uk) |>
  terra::mask(uk)
```

<!-- 
```{r}
#| label: virtual_srtm_plot
#| echo: false
#| fig-cap: 3 sec elevation data for UK obtained with {geodata} package.

terra::plot(vrt)
```
-->

Let's extract the elevation from our more accurate data sets:

```{r}
#| label: occ_elevation_2
occ <- occ |>
  dplyr::mutate(ele_30s = terra::extract(ele, terra::vect(geometry))$GBR_elv_msk) |>
  dplyr::mutate(ele_3s = terra::extract(vrt, terra::vect(geometry))$elevation_vrt)

occ |>
  head(5)
```

We got quite similar results regardless of resolution. On the other hand, if the your plots are located in mountainous region --- the higher the raster resolution, the better.

### TanDEM-X

TanDEM data is available from [https://download.geoservice.dlr.de/TDM90/](https://download.geoservice.dlr.de/TDM90/) service. To download the necessary data `TanDEM` package [@TanDEM] can be used, available on [GitHub](https://github.com/meteosimon/TanDEM-R).

<!--
Please remember to set up the keyring with your user name, like:

```{r}
#| eval: false

keyring::key_set(service = "geoservice.dlr", username = "gsapijaszko")
```

```{r}
#| label: tandem
#| eval: false
#| message: false

if(!dir.exists("data/tandem")) { dir.create("data/tandem") }
TanDEM::download_TanDEM(
  lon = c(bbox[1], bbox[3]),
  lat = c(bbox[2], bbox[4]),
  usr = "gsapijaszko",
  srv = "geoservice.dlr",
  dstdir = "data/tandem"
)
```

In next step we will build virtual raster and assign `NA` for cells with no data:
-->

```{r}
#| label: tandem_vrt
#| echo: false

r <- list.files(path = "data/tandem/", pattern = "TDM1_.+.tif", full.names = TRUE)
tandem <- terra::vrt(r, "data/tandem_vrt.vrt", overwrite = TRUE) 

terra::NAflag(tandem) <- -32767
```

```{r}
#| label: tandem_plot
#| eval: false
#| output: false
#| echo: false
#| 
tandem <- tandem |>
  terra::mask(uk) |>
  terra::crop(uk)
terra::plot(tandem)
```

<!--
Finally, we can extract elevations:
-->

```{r}
#| label: occ_elevation_3
#| warning: false
#| echo: false

occ <- occ |>
  dplyr::mutate(ele_tandem = terra::extract(tandem, terra::vect(occ))$tandem_vrt) 

```

### Copernicus data

The original EU-DEM v1.0 (or v1.1) data can be accessed from [copernicus.eu](https://dataspace.copernicus.eu/explore-data/data-collections/copernicus-contributing-missions/collections-description/COP-DEM) portal. It's divided in 10x10 degree tiles, which results in quite huge files (up to 5 GB per tile). The similar data set, in smaller tiles is available from [AWS Opendata Registry](https://registry.opendata.aws/copernicus-dem/). Details in [readme](https://copernicus-dem-30m.s3.amazonaws.com/readme.html). This set can be downloaded by `CopernicusDEM` package [@CopernicusDEM]. Please note, that this dataset is a Digital Surface Model (DSM) that represents the surface of the Earth including buildings, infrastructure and vegetation..

```{r}
#| label: copernicus
#| eval: false
#| echo: true
library(CopernicusDEM)

polygon <- sf::st_as_sf(uk)
sf::sf_use_s2(use_s2 = FALSE)

if(!dir.exists("data/copernicus")) { dir.create("data/copernicus")}

aoi_geom_save_tif_matches(
  sf_or_file = polygon,
  dir_save_tifs = "data/copernicus",
  resolution = 30,
  crs_value = 4326,
  threads = parallel::detectCores(),
  verbose = TRUE
)
```

```{r}
#| label: copernicus_vrl

r <- list.files(path = "data/copernicus/", pattern = "Copernicus_.+.tif", full.names = TRUE)
cop <- terra::vrt(r, "data/copernicus_vrt.vrt", overwrite = TRUE) 

occ <- occ |>
  dplyr::mutate(ele_cop = terra::extract(cop, terra::vect(geometry))$copernicus_vrt)

```

### ElevatR - the black horse

As the name suggests, `elevatr` is a package [@elevatr] to access elevation (or tiles) data from various sources. For global coverage (outside USA) it uses [AWS Terrain Tiles](https://registry.opendata.aws/terrain-tiles/) service. With help of `get_elev_point()` function obtaining the elevation is quite simple, it doesn't require to fetch and load in any raster tiles:

```{r}
#| label: elevatr
#| message: false
#| warning: false

occ <- occ |>
  dplyr::mutate(ele_aws = elevatr::get_elev_point(occ, src = "aws")$elevation) 
```

All values extracted (or obtained) from various sources are presented in @tbl-elevations. Please note the difference in values, especially TanDEM-X, which is elevation not terrain model.

```{r}
#| label: tbl-elevations
#| tbl-cap: Elevations obtained from different DEM/DSM models for locations of _Calystegia x pulchra_ in UK
#| echo: false

occ |>
  sf::st_drop_geometry() |>
  head(6) |>
  subset(select = c(1,3:7)) |>
  kableExtra::kable(row.names = FALSE,
                    col.names = c("key", "30s SRTM", "3s SRTM", "TanDEM-X", "Copernicus DEM", "AWS Terrain Tiles"),
                    escape = FALSE,
                    digits = 2,
                    format.args = list(decimal.mark = ".", big.mark = "")) |>
  kableExtra::kable_classic_2()

```


```{r}
#| label: write_bibliography
#| echo: false
#| include: false
#| eval: false
packages <- c("terra", "geodata", "rgbif", "CopernicusDEM", "TanDEM", "keyring", "elevatr")
knitr::write_bib(packages, file = "../../packages.bib")
```
